---
title: "Assignment_1_Data_624"
author: "Mubashira Qari"
date: "2026-02-01"
output: html_document
---

# 2.10 Exercises
# 1. Explore the following four time series: Bricks from aus_production, Lynx from pelt, Close from gafa_stock, Demand from vic_elec.

# Use ? (or help()) to find out about the data in each series.

# Bricks from aus_production

```{r}
library(fpp3)

# Access help documentation
?aus_production

```

# What is the time interval of each series?

Quarterly production of selected commodities in Australia.

Description
Quarterly estimates of selected indicators of manufacturing production in Australia.

Format
Time series of class tsibble.

Details
aus_production is a half-hourly tsibble with six values:

Beer:	Beer production in megalitres.
Tobacco:	Tobacco and cigarette production in tonnes.
Bricks:	Clay brick production in millions of bricks.
Cement:	Portland cement production in thousands of tonnes.
Electricity:	Electricity production in gigawatt hours.
Gas:	Gas production in petajoules.
Source
Australian Bureau of Statistics, catalogue number 8301.0.55.001 table 1.

# Use autoplot() to produce a time plot of each series.

```{r}
# Plot of the Bricks time series
aus_production |>
  autoplot(Bricks) +
  labs(title = "Australian Quarterly Clay Brick Production", y = "Millions of Bricks")
```
# Lynx from pelt

```{r}
# Access help documentation
?pelt

```

# What is the time interval of each series?
 
This is Annual Pelt trading records
Description
Hudson Bay Company trading records for Snowshoe Hare and Canadian Lynx furs from 1845 to 1935. This data contains trade records for all areas of the company.

Format
Time series of class tsibble

Details
pelt is an annual tsibble with two values:

Hare:	The number of Snowshoe Hare pelts traded.
Lynx:	The number of Canadian Lynx pelts traded.
Source
Hudson Bay Company

# Use autoplot() to produce a time plot of each series.

```{r}
# Plot of the Lynx time series
pelt |>
  autoplot(Lynx) +
  labs(title = "Canadian Annual Lynx Pelts Trade", y = "Number of Pelts")
```
# Close from gafa_stock

```{r}
# Access help documentation
?gafa_stock
```

# What is the time interval of each series?

This is Daily GAFA stock prices
Description
Historical stock prices from 2014-2018 for Google, Amazon, Facebook and Apple. All prices are in $USD.

Format
Time series of class tsibble

Details
gafa_stock is a tsibble containing data on irregular trading days:

Open:	The opening price for the stock.
High:	The stock's highest trading price.
Low:	The stock's lowest trading price.
Close:	The closing price for the stock.
Adj_Close:	The adjusted closing price for the stock.
Volume:	The amount of stock traded.
Each stock is uniquely identified by one key:

Symbol:	The ticker symbol for the stock.
Source
Yahoo Finance historical data

# Use autoplot() to produce a time plot of each series.

```{r}
# Plot of the Closing prices for all stocks
gafa_stock |>
  autoplot(Close) +
  labs(title = "GAFA Daily Closing Stock Prices", y = "Price ($US)")
```
# Demand from vic_elec

```{r}
# Access help documentation
?vic_elec
```

# What is the time interval of each series?

Half-hourly electricity demand for Victoria, Australia

Description
vic_elec is a half-hourly tsibble with three values:

Demand:	Total electricity demand in MWh.
Temperature:	Temperature of Melbourne (BOM site 086071).
Holiday:	Indicator for if that day is a public holiday.
Format
Time series of class tsibble.

Details
This data is for operational demand, which is the demand met by local scheduled generating units, semi-scheduled generating units, and non-scheduled intermittent generating units of aggregate capacity larger than 30 MWh, and by generation imports to the region. The operational demand excludes the demand met by non-scheduled non-intermittent generating units, non-scheduled intermittent generating units of aggregate capacity smaller than 30 MWh, exempt generation (e.g. rooftop solar, gas tri-generation, very small wind farms, etc), and demand of local scheduled loads. It also excludes some very large industrial users (such as mines or smelters).

Source
Australian Energy Market Operator.

# Use autoplot() to produce a time plot of each series.

```{r}
# Plot of the Electricity Demand
vic_elec |>
  autoplot(Demand) +
  labs(title = "Australia: Victoria Half-Hourly Electricity Demand", y = "MWh")
```

# For the last plot, modify the axis labels and title.

```{r}
# Plot of the Electricity Demand
vic_elec |>
  autoplot(Demand) +
labs(
    title = "Victoria Electricity Half-Hourly Demand",
    y = "Demand (Megawatts)",
    x = "Time (Half-Hourly)"
  )
```

# 2. Use filter() to find what days corresponded to the peak closing price for each of the four stocks in gafa_stock.

```{r}
view(gafa_stock)
```


```{r}
gafa_stock |>
  group_by(Symbol) |>
  filter(Close == max(Close)) |>
  arrange(desc(Close)) -> gafa_peak_closing_price

view(gafa_peak_closing_price)
```
# Download the file tute1.csv from the book website....

# a. You can read the data into R with the following script:

```{r message=FALSE}

tute1 <- readr::read_csv("https://raw.githubusercontent.com/uzmabb182/Data_624_Predictive_Analytics/refs/heads/main/tute1.csv")
View(tute1)
```

# b. Convert the data to time series

```{r}

mytimeseries <- tute1 |>
  mutate(Quarter = yearquarter(Quarter)) |>
  as_tsibble(index = Quarter)

view(mytimeseries)
```

# c. Construct time series plots of each of the three series

```{r}
mytimeseries |>
  pivot_longer(-Quarter) |>
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y")
```
# Check what happens when you don’t include facet_grid().

```{r}
mytimeseries |>
  pivot_longer(-Quarter) |>
  ggplot(aes(x = Quarter, y = value, colour = name)) +
  geom_line()
```
facet_grid with scales = "free_y" allow each variable to have its own y-axis range.
Without it all variables will fit on the same y-axis.
The small variable will look like a flat straight line at the bottom of the chart and the pattern will not be displayed clearly.

# The USgas package contains data on the demand for natural gas in the US.

# a. The USgas package contains data on the demand for natural gas in the US

```{r}
```

