---
title: "Assignment_4_Data_624"
author: "Mubashira Qari"
date: "2026-02-16"
output:
  html_document:
    toc: true
    toc_float: true
  word_document:
    toc: true
---

### Exercise 3.1 (a) Visualizing Distributions and Relationships

To understand the individual distributions of the nine predictors and the relationships between them, histograms and a correlation matrix are utilized.

```{r warning=FALSE}
library(mlbench)
library(tidyverse)
library(corrplot)

data(Glass)

# Visualizing Distributions (Histograms)
Glass %>%
  pivot_longer(cols = RI:Fe, names_to = "Predictor", values_to = "Value") %>%
  ggplot(aes(x = Value)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  facet_wrap(~ Predictor, scales = "free") +
  labs(title = "Distributions of Glass Predictors")

# Visualizing Relationships (Correlation Matrix)
cor_matrix <- cor(Glass[, 1:9])
corrplot(cor_matrix, method = "color", type = "upper", 
         addCoef.col = "black", number.cexs = 0.7)
```

### (b) Outliers and Skewness

Based on the visualizations and statistical summaries, the following observations are made regarding the data quality:

Skewness: Several predictors exhibit significant skewness. 
Elements like Ba (Barium), Fe (Iron), and K (Potassium) are highly right-skewed because many samples have values near zero.
RI (Refractive Index) and Ca (Calcium) also show moderate right skewness.

Outliers: The histograms and boxplots (if generated) reveal distinct outliers in nearly all predictors, 
particularly in K, Ba, and RI. For instance, K has a few samples with much higher values than the rest of the cluster.

```{r message=FALSE}
# Check skewness numerically
library(e1071)
apply(Glass[, 1:9], 2, skewness)
```

### (c) Relevant Transformations

To improve a classification model, specific transformations should be applied to address the issues identified in part (b):

Box-Cox Transformation: This is effective for resolving skewness in predictors like RI, Ca, and Al. Note that it requires strictly positive values, 
so predictors with zeros (like Ba, Fe, and K) would require a prior shift or a different transformation.

Centering and Scaling: Since predictors like Si (values ~70) and Fe (values ~0.1) are on vastly different scales, 
centering and scaling are necessary for many models (e.g., SVM or KNN) to ensure all features contribute equally.

Spatial Sign Transformation: Because significant outliers are present, the spatial sign transformation can be used to 
project predictor values onto a sphere, effectively minimizing the impact of extreme values.

Principal Component Analysis (PCA): The correlation plot shows a strong relationship between RI and Ca. 
PCA can be used to reduce this redundancy and address collinearity.

```{r message=FALSE}
# Example of applying pre-processing using caret
library(caret)
glass_prep <- preProcess(Glass[, 1:9], 
                         method = c("BoxCox", "center", "scale", "spatialSign"))
glass_transformed <- predict(glass_prep, Glass[, 1:9])

# glass_transformed
```

### 3. Exercise 3.2: Soybean Missing Value Analysis

Investigating  missing data patterns.

```{r}
library(mlbench)
library(tidyverse)
library(naniar) # Helpful for missing data visualization

data(Soybean)

# Identify the number of missing values per class
Soybean |>
  group_by(Class) |>
  summarise(
    total_obs = n(),
    missing_values = sum(is.na(across(everything()))),
    pct_missing = (missing_values / (total_obs * ncol(Soybean))) * 100
  ) |>
  arrange(desc(pct_missing))
```
**Summary of Findings for Soybean Data**

Class-Specific Missingness: Missing values are concentrated in specific disease categories like phytophthora-rot, herbicide-injury, and 2-4-d-injury.

Implication: Deleting these rows would eliminate almost all examples of these specific diseases from the dataset, making the model unable to recognize them.

Resolution Strategy: For this dataset, using imputation (like knnImpute) or treating "missing" as its own categorical level is more effective than removing data.

### (a) Investigate Frequency Distributions for Categorical Predictors

Investigate the frequency distributions for the categorical predictors. 
Are any of the distributions degenerate in the ways discussed earlier in this
chapter?

A categorical predictor is a variable with categories (e.g., Gender, Yes/No, Type A/B/C).

A distribution is degenerate when it causes modeling problems. Common issues:

Zero variance predictor: Only one category appears and this variable provides no predictive value
Near-zero variance predictor: One category dominates heavily. Model may struggle to learn from rare category.
Very sparse categories: Some categories appear only a few times.Can cause unstable estimates.

**Look at Frequency Tables**

```{r}
table(Soybean$date)

# for all predictors:
#lapply(Soybean, table)

```
**Check for Zero / Near-Zero Variance (Using caret)**

```{r}

library(caret)

nzv <- nearZeroVar(Soybean, saveMetrics = TRUE)
nzv


```
**Conclusion**

Frequency distributions were examined using contingency tables and near-zero variance diagnostics. 
No predictors exhibited zero variance. However, some predictors show moderate imbalance, although none meet strict near-zero variance thresholds. 
Therefore, no categorical predictors are completely degenerate.

### b. Investigate Missing Data (18%)

**Confirm Overall Missing Percentage**

```{r}

mean(is.na(Soybean)) * 100

```

**Missing Percentage Per Predictor**

```{r}
colMeans(is.na(Soybean)) * 100

# Sorting them
#sort(colMeans(is.na(Soybean)) * 100, decreasing = TRUE)

```

This shows which predictors are most missing.

Some predictors have very high missing rates and others have none

**Is Missingness Related to Class?**

Create missing indicators for highly missing variables.

```{r}

Soybean$hail_missing <- ifelse(is.na(Soybean$hail), 1, 0)

table(Soybean$hail_missing, Soybean$Class)


```

**test statistically**

```{r}
chisq.test(table(Soybean$hail_missing, Soybean$Class))

```

**Conclusion**

issingness Overview: The overall missingness is confirmed at approximately 18%, but the distribution is highly uneven across predictors.

Variable Variance: Your sorted list identifies that some predictors are significantly more prone to missing values than others, which could impact model stability.

Informative Missingness: The table() and Chi-square results indicate that missing values are heavily concentrated in specific disease classes (such as phytophthora-rot and herbicide-injury).

Modelling Impact: Since missingness is associated with the response class, the data is not "Missing Completely at Random" (MCAR). 
Deleting these observations would systematically remove nearly all examples of certain diseases, making imputation the necessary next step.


### c. Develop a strategy for handling missing data, either by eliminating predictors or imputation.

**Elimination Strategy (Filtering)**

```{r message=FALSE}
# Load the necessary library
library(caret)

# Identifying and remove Near-Zero Variance (NZV) predictors
# This removes uninformative variables that could crash certain models
nzv_metrics <- nearZeroVar(Soybean, saveMetrics = TRUE)
soy_filtered <- Soybean[, !nzv_metrics$nzv]

# Preparing for Imputation
# Since most predictors are factors, they must be converted to dummy variables 
# to allow for numerical imputation methods like KNN
soy_dummy_model <- dummyVars(Class ~ ., data = soy_filtered)
soy_numeric <- predict(soy_dummy_model, newdata = soy_filtered)

# Apply KNN Imputation
# This estimates missing values based on the 5 most similar observations
# knnImpute also centers and scales the data automatically
soy_preproc <- preProcess(soy_numeric, method = "knnImpute")
soy_final <- predict(soy_preproc, soy_numeric)

# Verify no missing values remain
sum(is.na(soy_final))
```

Now the data is now suitable for predictive modeling by addressing the issues identified during exploration.

Near-Zero Variance Filtering: Predictors with little to no variation were removed to prevent numerical errors and reduce model complexity.

Dummy Variable Encoding: The categorical predictors were converted into a numerical format, which is a requirement for distance-based imputation methods like KNN.

KNN Imputation: Instead of discarding observations, missing values were estimated using the most similar neighboring samples. 
This was critical because the missingness was associated with specific disease classes, and deletion would have biased the model.

Standardization: The data was automatically centered and scaled during the imputation process, ensuring that all predictors are on a comparable scale for the classification algorithm.

