---
title: "HW_5_mqari"
author: "Mubashira Qari"
date: "2026-02-23"
output: html_document
---




```{r}
library(fpp3)

algeria_economy <- global_economy |>
  filter(Country == "Algeria")
algeria_economy |>
  autoplot(Exports) +
  labs(y = "% of GDP", title = "Exports: Algeria")
```
The simplest of the exponentially smoothing methods is naturally called simple exponential smoothing (SES)16. This method is suitable for forecasting data with no clear trend or seasonal pattern.


```{r}

algeria_economy <- global_economy |>
  filter(Country == "Algeria")

algeria_fit <- algeria_economy |>
  model(ANN = ETS(Exports ~ error("A") + trend("N") + season("N")))

report(algeria_fit)
```

Take the mable, add the components of it and plot them

```{r}
components(algeria_fit) |> autoplot()
```
Now we left Join the fitted values to the components

```{r}
components(fit) |>
  left_join(fitted(fit), by = c("Country", ".model", "Year"))
```
Now generate the forecast
```{r}
algeria_fit |>
  forecast(h = "5 years") |>
  autoplot(algeria_economy) + 
 labs(y = "% of GDP", title = "Exports: Algeria")
```
Holt's Linear Trend

Component Form:
Forcast
Level
Trend

Example: Australian Population

Code to fit the model

```{r}
aus_economy <- global_economy |>
  filter(Code == "AUS")

aus_economy
```
After fitting the model it comes back with the estimates of the parameters

```{r}
aus_economy <- global_economy |>
  filter(Code == "AUS") |>
  mutate(Pop = Population/1e6)
  
aus_fit <- aus_economy |>
  model(ANN = ETS(Pop ~ error("A") + trend("A") + season("N")))

report(aus_fit)
```

If we take that model fit and add to component function, we get the individual component

```{r}
components(aus_fit) |> autoplot()
```
Now take the Components and add it to the fitted values
Data = Pop, Alpha level changes = l, slope changes = beta, remainder = fitted-Pop

```{r}
components(aus_fit) |> 
  left_join(fitted(aus_fit), by = c("Country", ".model", "Year"))
```

Visualizing Forecast Equation
It picked up the trend of the end of the series
```{r}
aus_fit |>
  forecast(h=10) |>
  autoplot(aus_economy) +
  labs(y = "Millions", title = "Poulation: Australia")
```
Damped Trend Method

It damped the trend
larger value of damping less damping occurs
smaller value then flattens quickly


```{r}
aus_economy <- global_economy |>
  filter(Code == "AUS") |>
  mutate(Pop = Population/1e6)
  
aus_economy |>
  model(holt = ETS(Pop ~ error("A") + trend("Ad") + season("N"))) |> # Added pipe here
  forecast(h = 20) |>
  autoplot(aus_economy)
```
Lets now see all the three models that we are introduced in this chapter

```{r}
aus_fit <- aus_economy |>
  filter(Year <=2010) |>
  model(
    SES = ETS(Pop ~ error("A") + trend("N") + season("N")),
    Holt = ETS(Pop ~ error("A") + trend("A") + season("N")),
    Damped = ETS(Pop ~ error("A") + trend("Ad") + season("N"))
  )
report(aus_fit)
```
SES: Simple Exponential Smoothing Model - Additive Error No Trend No Seasonality
Holt: Holt Linear Method - Additive Error Additive Trend and No Seasonality
Damped: Damped Trend - Additive Error Additive Damped trend and No Seasonality

tidy() function - get the coefficients
accuracy() function - get various measure of accuracy

```{r}
tidy(aus_fit) # esimates
accuracy(aus_fit) # accuracy metrics
```
This shows how to compare trends in the three models

We will cover how to compare seasonality in other sections below

Holt and Winters method captured seasonality

gamma = 0 no change

gamma = 1 close to seasonal naive

usually gamma <= 0.2

Using Additive Halt Winter Method and Multiplicative Halt Winter Method

```{r}
aus_holidays <- tourism |>
  filter(Purpose == "Holiday") |>
  summarize(Trips = sum(Trips))

holidays_fit <- aus_holidays |>
  model(
    additive = ETS(Trips ~ error("A") + trend("A") + season("A")),
    multiplicative = ETS(Trips ~ error("M") + trend("A") + season("M"))
  ) 
fc <- holidays_fit |> forecast()

report(holidays_fit)
```
Get the Estimated Coefficients

```{r}
tidy(holidays_fit) |>
  spread(.model, estimate)
```

Finding Estimated Components and left joining fitted values to have comparison

Additive: level+slope+season = fitted value
Multiplicative: level+slope*season = fitted value

```{r}
components(holidays_fit) |>
  filter(.model == "additive") |>
  left_join(fitted(holidays_fit), by = c(".model", "Quarter"))
```
```{r}
# Compare additive and multiplicative components side-by-side
holidays_fit |>
  components() |>
  left_join(fitted(holidays_fit), by = c(".model", "Quarter"))
```


```{r}
components(holidays_fit) |> autoplot()
```

```{r}
# 1. Fit both models
holidays_fit <- aus_holidays |>
  model(
    additive = ETS(Trips ~ error("A") + trend("A") + season("A")),
    multiplicative = ETS(Trips ~ error("M") + trend("A") + season("M"))
  )

# 2. Plot with .model assigned ONLY to columns
holidays_fit |>
  components() |>
  autoplot() +
  facet_grid(cols = vars(.model), scales = "free_y") + 
  labs(title = "Side-by-Side Comparison of ETS Components")
```
```{r}
library(tidyverse)

library(tidyverse)

holidays_fit |>
  components() |>
  # Ensure we are selecting the right columns
  pivot_longer(c(Trips, level, slope, season, remainder), 
               names_to = "component", values_to = "value") |>
  mutate(component = factor(component, 
         levels = c("Trips", "level", "slope", "season", "remainder"))) |>
  ggplot(aes(x = Quarter, y = value)) +
  geom_line() +
  # This is the key: independent scales for every single box
  facet_grid(rows = vars(component), cols = vars(.model), scales = "free_y") +
  theme_minimal() +
  labs(title = "Additive vs Multiplicative Components",
       subtitle = "Note: Multiplicative season/remainder are ratios",
       y = NULL)
```
Forecast using two methods (Additive & Multiplicative)

Multiplicative is better in adding seasonality here as it shows wider

```{r}
fc |>
  autoplot(aus_holidays, level =NULL) +
  labs(y = "Thousand", title = "Overnight trips")
```

Example: Holt-Winters method with daily data

The Holt-Winters method can also be used for daily type of data, where the seasonal period is  
m=7 . Here we forecast pedestrian traffic at a busy Melbourne train station in July 2016.

```{r}
pedestrian_per_day <- pedestrian |>
  filter(Sensor == "Southern Cross Station", yearmonth(Date) == yearmonth("2016 July")) |>
  index_by(Date) |>
  summarise(Count = sum(Count))

pedestrian_fit <- pedestrian_per_day |>
  model(ETS(Count ~ error("A") + trend("Ad") + season("M")))

pedestrian_fit %>% report()
```

Here we estimate 9 inital values, 1 for level, 1 for slope, and 7 for seasonal index.
```{r}
pedestrian_fit |>
  forecast(h = "2 weeks") |>
  autoplot(pedestrian_per_day)
```
Daily data with bust on weekends and peaks on weekdays

```{r}
sth_cross_ped <- pedestrian |>
  filter(
    Sensor == "Southern Cross Station", 
    Date >= "2016-07-01") |>
  index_by(Date) |>
  summarise(Count = sum(Count)/1000)

# Forecast and Plot
sth_cross_ped |>
  filter(Date <= "2016-07-31") |>
  model(ets = ETS(Count ~ error("M") + trend("Ad") + season("M"))) |>
  forecast(h = "2 weeks") |>
  autoplot(sth_cross_ped |> filter(Date <= "2016-08-14")) + # Use + for ggplot layers
  labs(
    title = "Daily traffic: Southern Cross",
    y = "Pedestrians ('000)"
  )
```

Example: Domestic holiday tourist visitor nights in Australia
Forcasting aus_hodidays tibble time series using ETS model
Model return Model: ETS(M,N,M) = Multiplicative Errors No Trend Multiplicative Seasonality

```{r}
aus_holidays <- tourism |>
  filter(Purpose == "Holiday") |>
  summarise(Trips = sum(Trips))  

holidays_fit <- aus_holidays |> 
  model(ETS(Trips))

holidays_fit |> report()
```
lets print out the components
here remainder is a relative error
```{r}
holidays_fit |>
  components() |> 
  autoplot() +
  labs(title = "ETS(M,N,M) components")
```
lets plot the residuals

```{r}
residuals(holidays_fit) |>
  autoplot() + 
  ggtitle("Innovation errors")

residuals(holidays_fit, type = "response") |> 
  autoplot() + 
  ggtitle("Response errors")
```
We can see the Innovation and residual  to list

```{r}
holidays_fit |>
  augment()
```

Forecasting

```{r}
holidays_fit |>
  forecast(h = 8) |> 
  autoplot(aus_holidays, level = 95)
```
Forecasting with ETS models

ETS Forecasting Workflow

```{r}
library(fpp3)

# 1. Prepare the data (Corticosteroid drug sales)
h02_data <- PBS |>
  filter(ATC2 == "H02") |>
  summarise(Cost = sum(Cost))

# 2. Fit the ETS model (Letting R choose the best model automatically)
h02_fit <- h02_data |>
  model(auto = ETS(Cost))
report(h02_fit)

# OR Choosing AAA Model manually - giving worse forecast seeing AIC value
h02_fit1 <- h02_data |>
  model(AAA = ETS(Cost ~  error("A") + trend("A") + season("A"))) |>
report(h02_fit1)

# 3. Generate forecasts for the next 2 years (24 months)
h02_fc <- h02_fit |>
  forecast(h = "2 years")

# 4. Plot the results
h02_fc |>
  autoplot(h02_data) +
  labs(title = "Corticosteroid Drug Sales Forecast",
       y = "$ Cost", x = "Month")
```


```{r}

tidy(h02_fit) # esimates
accuracy(h02_fit) # accuracy metrics

tidy(h02_fit1) # esimates
accuracy(h02_fit1) # accuracy metrics
```
Combining two two compare results

```{r}
h02_data <- PBS |>
  model(
    auto = ETS(Cost), 
    AAA = ETS(Cost ~  error("A") + trend("A") + season("A"))
  ) |>
  accuracy()
``` 



