---
title: "home_work_2_mq"
author: "Mubashira Qari"
date: "2026-02-08"
output: html_document
---

# Prepare data for decomposition 
# we make transformation and adjustment
# We devide the data by population to make it totals per capita


```{r}
library(fpp3)
global_economy

```


```{r}
global_economy |>
  filter(Country == "Australia") |>
autoplot(GDP)
```
# Per Capita Adjustment
To understand the growth of the economy it should be relative to the size of the population
for that we divide by population and then model that per capita data
This is what it means adjustment


```{r}
global_economy |>
  filter(Country == "Australia") |>
  autoplot(GDP/Population) +
  labs(title= "GDP per capita", y = "$US")
```

# Data Preparation (Tidy Data)

```{r}
gdppc <- global_economy |>
  mutate(GDP_per_capita = GDP/Population) |>  # interested in GDP_per_capita
  select(Year, Country, GDP, Population, GDP_per_capita)
gdppc
```

# Visulaize this data of GDP_per_capita for Country = Sweden

```{r}
gdppc |>
  filter(Country == "Sweden") |>
  autoplot(GDP_per_capita) +
  labs(title = "GDP per capita for Sweden", y="$US")
```
Its a yearly data, so there is no seasonal component. 
It has upward trend

# Model Specification & Estimation

Model function captures the time trend
Each country has a row for time series linear model.
Piping a table into a model gives mable

```{r}
fit <- gdppc |>
  model(trend_model = TSLM(GDP_per_capita ~ trend()))

fit
```
We will skip model evaluation step for later but now we will generate forecast

# Generating Forecast

A fable is a forecast table with point forecasts and distributions.
For each county we get three forecast since h=3
GDP_per_capita column contains distribution and point forecast is a point from that distribution which usually is the mean in mean column
A forecast distribution is normal with mean and variance

```{r}
fit |> forecast(h = "3 years")

```

# Now Visualizing and Evaluating Forecast

```{r}
fit |> 
  forecast(h="3 years") |>
  filter(Country =="Sweden") |>
  autoplot(gdppc) + labs(title = "GDP per capita for Sweden", y = "$US")
```
# Four Simple Methods of Forcasting for Comparision

# Mean(y): Average Method

Y1 - YT = Historical Data
Forecast: Y^(T+h)/T = Ybar = (Y1 - YT)/T = Average of historical data



# NAIVE(y): Naive method (best for financial data)

Naive method = Last observed value of the data becomes the value of your forecast.
Your point forecast for all future periods

YT = Last Historical Data Point
Forecast: Y^(T+h)/T = YT


# SNAIVE(y ~lag(m)): Seasonal Naive Method

If you are trying to forecast Quarter 1, then you take the last time you observed Quarter 1

If you are trying to forecast Quarter 2, then you take the last time you observed Quarter 2

So basically, the last full year of data gets replicated to the future foecasts

And this gives us the future point forecast

# RW(y ~ drift()): Drift Method

Forecast = last observed value (YT) + Average Change in data over time

This is equivalent to extrapolating a line drawn between first and last observations

```{r}
# The filter_index() function is a convenient shorthand for extracting a section of a time series.

bricks <- aus_production |>
  filter_index("1970 Q1" ~ "2004 Q4") |>
  select(Bricks)
```

```{r}
bricks |> model(MEAN(Bricks))
bricks |> model(NAIVE(Bricks))
bricks |> model(SNAIVE(Bricks ~ lag("year")))
bricks |> model(RW(Bricks ~ drift()))
```

# Model Fitting

The model() function trains models to data


```{r}
library(fpp3)

# 1. Ensure aus_production is the starting point (the data)
brick_fit <- aus_production |>
  filter(!is.na(Bricks)) |>
  # 2. Apply the model definitions INSIDE the model() function
  model(
    Seasonal_naive = SNAIVE(Bricks ~ lag("year")),
    Naive          = NAIVE(Bricks),
    Drift          = RW(Bricks ~ drift()),
    Mean           = MEAN(Bricks) # Note: Use MEAN(Bricks) directly
  )

# To see the results (the mable)
brick_fit
```

# Producing Forecast (Point Forecast)

```{r}
brick_fc <- brick_fit |>
  forecast(h="5 years")
brick_fc
```

# Visulasing Forecasts

```{r}
brick_fc |>
  autoplot(aus_production, level = NULL) +
  labs(title = "Clay brick production in Australia",
       y ="Millions of bricks") +
  guides(colour = guide_legend(title="Forecast"))
```
# Fitted values
Each observation in a time series can be forecast using all previous observations. We call these fitted values.
fitted values are the forecast or predictions.

# Residuals
The “residuals” in a time series model are what is left over after fitting a model. 

# Residuals in forecasting: 
Difference between observed value and fitted value

# Assumptions for Residuals:
Residuals are uncorrelated. If they are not, then the information left in residuals should be used in computing forecasts.
Residuals have mean zero. I they don't then forecasts are biased.

Useful Properties: 
Residuals have constant variance.
residuals are normally distributed

Lets take an example:

```{r}
fb_stock
```


```{r}
fb_stock |> autoplot(Close)
```
# Fit the Model (NAIVE)
augment function give us fitted values, residuals,and innovation residuals column
When Testing model we always use innov column
first value is missing since we don't know the value before that observation

```{r}
fb_stock <- google_stock |>
  mutate(trading_day = row_number()) |>
  as_tsibble(index = trading_day, regular = TRUE)
fb_stock
```

```{r}
fit <- fb_stock |>
  model(NAIVE(Close))
augment(fit)
```
Stock market data only exists for business days. If you plot stock prices using a standard date column, your graph will show literal gaps or long flat lines over weekends and holidays. This "breaks" many time series models that require a regular, consecutive index.


# Plotting the Observations and the Fitted values

```{r}
augment(fit) |>
  ggplot(aes(x = trading_day)) +
  geom_line(aes(y = Close, colour = "Data")) +
  geom_line(aes(y = .fitted, colour = "Fitted")) 
```
# Creating vertical Differences


```{r}
augment(fit) |>
  filter(trading_day > 900) |>
  ggplot(aes(x = trading_day)) +
  geom_line(aes(y = Close, colour = "Data")) +
  geom_line(aes(y = .fitted, colour = "Fitted")) 
```
# Plotting the Residuals from NAIVE method (Time Plot)

you can see they are centered around zero
But few big spikes
Also more variation in the end then rest of the plot
This all suggest that our assumptions are not quite right here

```{r}
augment(fit) |>
  autoplot(.resid) +
  labs(y = "$US",
       title = "Residuals from naive method")
```
# Histogram of Residuals

```{r}
augment(fit) |>
 ggplot(aes(x = .resid)) +
  geom_histogram(bins =150) +
  labs(title = "Histogram of Residuals")
```
It's not quite normal since we see outliers around and shape not quite normal

# Plot ACF of Residuals

```{r}
augment(fit) |>
 ACF(.resid) |>
  autoplot() +
  labs(title = "ACF of Residuals")
```

Lines outside the blue line is not right

# Plot all residual plots togather

```{r warning==FALSE}
gg_tsresiduals(fit)
```
# Portmanteau tests for autocorrelation for Residuals

The Ljung-Box test is a formal statistical test used to determine if the residuals of your model are "white noise" (random) or if there is still a pattern (autocorrelation) left behind that your model failed to capture.

1. The Key Result: The P-Value
In statistics, we typically compare the p-value to a significance level (usually alpha = 0.05):If p > 0.05: The residuals are indistinguishable from white noise (This is good!).If p < 0.05: The residuals have significant autocorrelation (The model missed something).

2. What this means for your NAIVE model
The fact that your p-value is significant means there is information left in the residuals that the NAIVE model is not using.

Specifically, it suggests that "yesterday's error" is correlated with "today's error." Even though the NAIVE model is a common benchmark for stocks, this test tells us that it hasn't perfectly captured the underlying dynamics of the Google stock price.

3. Visualizing the "Information"
ACF (Autocorrelation Function) plot of these residuals, we see one or more bars poking out beyond the blue dashed significance lines.

Conclusion: Since 0.0005349 is much smaller than 0.05, we reject the null hypothesis. Your residuals are not white noise.

```{r}
augment(fit) |>
  features(.resid, ljung_box, lag = 10)
```

# Probability Distribution of the Forecast

Assuming residual are normal, uncorrelated, and Std Deviation is constant (Homoscadastic) 
then we can derive the Probability distribution for the four benchmark methods

# Forecast Distribution - Pridiction Intervals

```{r}
aus_production |>
  filter(!is.na(Bricks)) |>
  model(Seasonal_naive = SNAIVE(Bricks)) |>
  forecast(h = "5 years")
```

# To get the Hi and low limit of Pridiction Intervals
use hilo function (95 % Interval)
```{r}
aus_production |>
  filter(!is.na(Bricks)) |>
  model(Seasonal_naive = SNAIVE(Bricks)) |>
  forecast(h = "5 years") |>
  hilo(level =95)
```


```{r}
aus_production |>
  filter(!is.na(Bricks)) |>
  model(Seasonal_naive = SNAIVE(Bricks)) |>
  forecast(h = "5 years") |>
  hilo(level =95) |>
  mutate(lower = `95%`$lower, upper = `95%`$upper)
```




# Example: Australian quarterly beer production

first three methods applied to Australian quarterly beer production from 1992 to 2006, with the forecasts compared against actual values in the next 3.5 years.

```{r}
# Set training data from 1992 to 2006
train <- aus_production |>
  filter_index("1992 Q1" ~ "2006 Q4")
# Fit the models
beer_fit <- train |>
  model(
    Mean = MEAN(Beer),
    `Naïve` = NAIVE(Beer),
    `Seasonal naïve` = SNAIVE(Beer)
  )
# Generate forecasts for 14 quarters
beer_fc <- beer_fit |> forecast(h = 14)
# Plot forecasts against actual values
beer_fc |>
  autoplot(train, level = NULL) +
  autolayer(
    filter_index(aus_production, "2007 Q1" ~ .),
    colour = "black"
  ) +
  labs(
    y = "Megalitres",
    title = "Forecasts for quarterly beer production"
  ) +
  guides(colour = guide_legend(title = "Forecast"))
```

# Example: Google’s daily closing stock price

```{r}
# Re-index based on trading days
google_stock <- gafa_stock |>
  filter(Symbol == "GOOG", year(Date) >= 2015) |>
  mutate(day = row_number()) |>
  update_tsibble(index = day, regular = TRUE)
# Filter the year of interest
google_2015 <- google_stock |> filter(year(Date) == 2015)
# Fit the models
google_fit <- google_2015 |>
  model(
    Mean = MEAN(Close),
    `Naïve` = NAIVE(Close),
    Drift = NAIVE(Close ~ drift())
  )
# Produce forecasts for the trading days in January 2016
google_jan_2016 <- google_stock |>
  filter(yearmonth(Date) == yearmonth("2016 Jan"))
google_fc <- google_fit |>
  forecast(new_data = google_jan_2016)
# Plot the forecasts
google_fc |>
  autoplot(google_2015, level = NULL) +
  autolayer(google_jan_2016, Close, colour = "black") +
  labs(y = "$US",
       title = "Google daily closing stock prices",
       subtitle = "(Jan 2015 - Jan 2016)") +
  guides(colour = guide_legend(title = "Forecast"))
```

#=============================================
# Inflation Adjustment + Calendar Adjustment

```{r}
aus_retail
```

```{r}
print_retail <- aus_retail |>
  filter(Industry == "Newspaper and book retailing") |>
  group_by(Industry) |>
  index_by(Year = year(Month)) |>
  summarise(Turnover = sum(Turnover))

print_retail
```

```{r}
print_retail |> autoplot(Turnover)
```

```{r}

aus_economy <- global_economy |>
  filter(Code == "AUS")

aus_economy
```


```{r}

print_retail |>
  left_join(aus_economy, by = "Year")

```


```{r}
print_retail |>
  left_join(aus_economy, by = "Year") |>
  mutate(Adjusted_turnover = Turnover / CPI * 100) |>
  pivot_longer(c(Turnover, Adjusted_turnover),
               values_to = "Turnover") |>
  mutate(name = factor(name,
         levels=c("Turnover","Adjusted_turnover"))) |>
  ggplot(aes(x = Year, y = Turnover)) +
  geom_line() +
  facet_grid(name ~ ., scales = "free_y") +
  labs(title = "Turnover: Australian print media industry",
       y = "$AU")
```

# Mathematical Adjustment
If the data show different variation at different levels of the series, then a transformation can be useful


```{r}
food <-aus_retail |>
  filter(Industry ==  "Food retailing") |>
  summarise(Turnover = sum(Turnover)) 

food
```
Idea of transformation is to make graph more homogeneous (more even variation across the scale of data).

```{r}
# 2. Plot the data
food |>
  autoplot(Turnover) +
  labs(
    title = "Total Australian Food Retailing Turnover",
    y = "Turnover (million $AUD)",
    x = "Month"
  )
```


```{r}
# litter more even variation 
food |> autoplot(sqrt(Turnover))+
  labs(y = "Square root turnover")

```
```{r}
# litter more even variation with cube root
food |> autoplot(Turnover^(1/3))+
  labs(y = "Cube root turnover")
```
```{r}
#  even variation with log root, remove heterogeneity so you fit more simple model
food |> autoplot(log(Turnover)) +
  labs(y = "Log root turnover")
```
```{r}
# this is taking too far with more variation in the beginning and less on the end. So log is best solution
food |> autoplot(-1/Turnover) +
  labs(y = "Inverse turnover")
```

# Box Cox Transformation
Each of these transformations can be thought of in a continuous scale of transformations which are known as Box Cox transformations. 
Lamda describe how strong transformation is
If Lambda is 0, we take natural logrithm
If Lambda is not equal to 0, we do this tranformation which raises the data to the power of Lambda and then subracts 1/lambda 
Reason for sign is to be able to deal with negative value - called pickle docson tranformation

When we start making lambda smaller, transformation gets stronger and we choose where transformation is even
It remove heterogeneity in the data

```{r}
lambda <- aus_production |>
  features(Gas, features = guerrero) |>
  pull(lambda_guerrero)
aus_production |>
  autoplot(box_cox(Gas, lambda)) +
  labs(y = "",
       title = latex2exp::TeX(paste0(
         "Transformed gas production with $\\lambda$ = ",
         round(lambda,2))))
```

There is function that automatically choose lambda called guerrero transformation and gives you most even distribution across the scale

```{r}
food |>
  features(Turnover, features =guerrero)
```
Transformation have large impact on predictions

another log 1P tranformation when data have zeros but you want something like log so it adds 1 and then takes a log.

Choosing log has nice effect of forcing all forecast to stay positive

When you work on the tranform data, then you have to back tranform to see the results on the original scale.

Using Fable package, we don't have to undo transformations when doing forcasting later on since will be done automatically


#Now Prepare data for decomposition 
Split the data into these components
Look for Trend, Cyclic, Seasonal

Trend: pattern exist when there is long tem increase or decrease in the data
Cycle: pattern exist when data exhibit rise and fall that are not for fixed period
Seasonal: pattern exist when a series is influenced by seasonal factor (eg,quarter, month,day of week)

In time series decomposition we split the data into following

y = f(S, T, R)

y =data at period t
T= trend cycle component at t (combine trend and cycle)
S= Sesonal cycle component at t
R = remainder cycle component at t

and it joined togather in one of two ways:

Additive Decomposition: y =S+T+R
Multiplicative Decomposition: y = SX T X R

```{r}
us_employment
```

```{r}
us_retail_employment <- us_employment |>
  filter(year(Month) >= 1990, Title == "Retail Trade") |>
  select(-Series_ID)
autoplot(us_retail_employment, Employed) +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")
```

```{r}
# Seasonal Trend and lowest decomposition
dcmp <- us_retail_employment |>
  model(stl = STL(Employed))
components(dcmp)

```

Plot the data
```{r}
components(dcmp) |> autoplot()
```

```{r}
components(dcmp) |>
  as_tsibble() |>
  autoplot(Employed, colour="gray") +
  geom_line(aes(y=trend), colour = "#D55E00") +
  labs(
    y = "Persons (thousands)",
    title = "Total employment in US retail"
  )
```
Sub Series Plot of seasonal component: show the shape of seasonality

```{r}
components(dcmp) |> gg_subseries(season_year)
```

Seasonally Adjusted Data: Remove the seasonality
Helps see fluctuations in the data that are not affected by the seasonal pattern
most commonly used in the decomposition

Additive Decomposition: y-S = T+R

Mutiplicative Decomposition: y/S = T X R

```{r}
components(dcmp) |>
  as_tsibble() |>
  autoplot(Employed, colour = "gray") +
  geom_line(aes(y=season_adjust), colour = "#0072B2") +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")
```

If the seasonal component is removed from the original data, the resulting values are the “seasonally adjusted” data.

For an additive decomposition, the seasonally adjusted data are given by  
y-S
 , and for multiplicative data, the seasonally adjusted values are obtained using  
y/S

Its more wiggly because you remove the seasonality and what left is trend and remainder

it is better to use trend-cycle component to look for turning point

# Classical Decomposition

for performing classical decomposition, first thing to look is the trend component and its estimate
and for estimate of trend component we look into moving averages

```{r}
global_economy |>
  filter(Country == "Australia") |>
  autoplot(Exports) +
  labs(y = "% of GDP", title = "Total Australian exports")
```

disadvantage: Can not be estimated near the end points. we loose point at the end

to overcome this

This is easily computed using slide_dbl() from the slider package which applies a function to “sliding” time windows. In this case, we use the mean() function with a window of size 5.

```{r}
aus_exports <- global_economy |>
  filter(Country == "Australia") |>
  mutate(
    `5-MA` = slider::slide_dbl(Exports, mean,
                .before = 2, .after = 2, .complete = TRUE)
  )
```

To see what the trend-cycle estimate looks like, we plot it along with the original data in Figure

```{r}
aus_exports |>
  autoplot(Exports) +
  geom_line(aes(y = `5-MA`), colour = "#D55E00") +
  labs(y = "% of GDP",
       title = "Total Australian exports")
```
Moving averages of averages

```{r}
beer <- aus_production |>
  filter(year(Quarter) >= 1992) |>
  select(Quarter, Beer)
beer_ma <- beer |>
  mutate(
    `4-MA` = slider::slide_dbl(Beer, mean,
                .before = 1, .after = 2, .complete = TRUE),
    `2x4-MA` = slider::slide_dbl(`4-MA`, mean,
                .before = 1, .after = 0, .complete = TRUE)
  )
```

Example: Employment in the US retail sector

```{r}
us_retail_employment_ma <- us_retail_employment |>
  mutate(
    `12-MA` = slider::slide_dbl(Employed, mean,
                .before = 5, .after = 6, .complete = TRUE),
    `2x12-MA` = slider::slide_dbl(`12-MA`, mean,
                .before = 1, .after = 0, .complete = TRUE)
  )
us_retail_employment_ma |>
  autoplot(Employed, colour = "gray") +
  geom_line(aes(y = `2x12-MA`), colour = "#D55E00") +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail")
```
Weighted moving averages
Combinations of moving averages result in weighted moving averages

# Classical Decomposition

 Two types of Classical decomposition
 
 Additive: where components add
 Mutiplicative: where components mutiply
 
 First step is to use moving averages to get trend so I can estimate
 
 Estimate T Trend using (2xm)  Moving Average if m is even
 otherwise T using m Moving Average
 
 Next Calculate de-trended series
 
 Additive decomposition: y-T
 Mutiplicative decomposition: y/T
 
 De-trending remove smoothed series T from and leave S and R
 
```{r}
us_retail_employment |>
  model(
    classical_decomposition(Employed, type = "additive")
  ) |>
  components() |>
  autoplot() +
  labs(title = "Classical additive decomposition of total
                  US retail employment")
```
 
X-11 Decomposition Method:

```{r}
x11_dcmp |>
  ggplot(aes(x = Month)) +
  geom_line(aes(y = Employed, colour = "Data")) +
  geom_line(aes(y = season_adjust,
                colour = "Seasonally Adjusted")) +
  geom_line(aes(y = trend, colour = "Trend")) +
  labs(y = "Persons (thousands)",
       title = "Total employment in US retail") +
  scale_colour_manual(
    values = c("gray", "#0072B2", "#D55E00"),
    breaks = c("Data", "Seasonally Adjusted", "Trend")
  )
```

